---
title: "emulate-jules-all-timeseries"
author: "Doug McNeall"
date: "11/7/2023"
output: html_document
---


# Introduction
This notebook build a set high-dimensional emulators of timeseries of JULES global mean carbon cycle outputs. It uses PCA to reduce the dimension of the output, and build a Gaussian Process emulator of the output in reduced dimension.  

This notebook builds on the work of explore-jules-timeseries-emulate, and applies the emulator to multiple timeseries.


## Preliminaries
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
# Libraries
library(DiceKriging)
library(parallel)
```


```{r}
# Helper functions

anomalizeTSmatrix = function(x, ix){
  # Anomalise a timeseries matrix
  subx = x[ ,ix]
  sweepstats = apply(subx, 1, FUN=mean)
  anom = sweep(x, 1, sweepstats, FUN = '-')
  anom
}

mat2list <- function(X){
  
  # Turns the p columns of a matrix into a p length list,
  # with each column becoming an element of the list
  
  out <- vector(mode = 'list', length = ncol(X))
  for(i in 1:ncol(X)){
    out[[i]] <- X[ , i]
  }
  out
}


reset <- function() {
  # Allows annotation of graphs, resets axes
  par(mfrow=c(1, 1), oma=rep(0, 4), mar=rep(0, 4), new=TRUE)
  plot(0:1, 0:1, type="n", xlab="", ylab="", axes=FALSE)
}

makeTransparent<-function(someColor, alpha=100)
  # Transparent colours for plotting
{
  newColor<-col2rgb(someColor)
  apply(newColor, 2, function(curcoldata){rgb(red=curcoldata[1], green=curcoldata[2],
                                              blue=curcoldata[3],alpha=alpha, maxColorValue=255)})
}



pca_km <- function(X, Y, train_ix, test_ix, npctol = 0.99, scale = FALSE, center = TRUE, ...){
  # emulate high dimensional output  
  
  require(parallel)
  
  # Split into training and test samples
  X_train <- X[train_ix, ]
  X_test  <- X[test_ix, ]
  
  Y_train <- Y[train_ix, ]
  Y_test  <- Y[test_ix, ]
  
  
  #reduce dimension of the output
  pca <- prcomp(Y_train, scale = scale, center = center)
  
  
  # choose a number of pcs
  pca_summary <- summary(pca)
  
  # 2 PCs minimum
  if(pca_summary$importance[3,2] < npctol){
    
  npc <- as.numeric(which(pca_summary$importance[3,] > npctol)[1])
  
  }
  
  else{
    npc <- 2
    }
  
  
  scores_train_list <- mat2list(pca$x[, 1:npc])
  
  # fitting the emulator is a slow step, so we use parallel computation
  # Fit an emulator to each principal component in turn
  fit_list <- mclapply(X = scores_train_list, FUN = km, formula = ~., design = X_train,
                       mc.cores = 4, control = list(trace = FALSE, maxit = 200))
  
  scores_em_test_mean  <- NULL
  scores_em_test_sd    <- NULL
  
  scores_em_train_mean <- NULL
  scores_em_train_sd   <- NULL
  
  for(i in 1:npc){
    
    loo <- leaveOneOut.km(fit_list[[i]], type = 'UK', trend.reestim = TRUE)
    pred_test <- predict(fit_list[[i]], newdata = X_test, type = 'UK')
    
    # Predict training data (low dimension representation)
    scores_em_train_mean <- cbind(scores_em_train_mean, loo$mean)
    scores_em_train_sd <- cbind(scores_em_train_sd, loo$sd)                            
    
    # Predict test data (low dimension representation)                         
    scores_em_test_mean <- cbind(scores_em_test_mean, pred_test$mean)
    scores_em_test_sd   <- cbind(scores_em_test_sd, pred_test$sd)
    
  }
  
  # Project back up to high dimension
  if(scale){
    anom_train <- pca$rotation[ ,1:npc] %*% t(scores_em_train_mean[,1:npc])*pca$scale
    anom_test <- pca$rotation[ ,1:npc] %*% t(scores_em_test_mean[,1:npc])*pca$scale
    
    sd_train <- t(pca$rotation[ ,1:npc] %*% t(scores_em_train_sd[,1:npc])*pca$scale)
    sd_test <- t(pca$rotation[ ,1:npc] %*% t(scores_em_test_sd[,1:npc])*pca$scale)
  }
  
  else{
    anom_train <- pca$rotation[ ,1:npc] %*% t(scores_em_train_mean[,1:npc])
    anom_test <- pca$rotation[ ,1:npc] %*% t(scores_em_test_mean[,1:npc])
    
    sd_train <- t(pca$rotation[ ,1:npc] %*% t(scores_em_train_sd[,1:npc]))
    sd_test <- t(pca$rotation[ ,1:npc] %*% t(scores_em_test_sd[,1:npc]))
    
  }
  
  Ypred_train <- t(anom_train + pca$center)
  Ypred_test <- t(anom_test + pca$center)
  
  Yerr_train <- Y_train - Ypred_train
  Yerr_test <- Y_test - Ypred_test
  
  return(list(X = X,
              Y = Y,
              train_ix = train_ix,
              test_ix = test_ix,
              X_train = X_train,
              X_test = X_test,
              npc = npc,
              pca = pca,
              fit_list = fit_list,
              scores_em_train_mean = scores_em_train_mean,
              scores_em_train_sd = scores_em_train_sd,
              scores_em_test_mean = scores_em_test_mean,
              scores_em_test_sd = scores_em_test_sd,
              Ypred_train = Ypred_train,
              Ypred_test = Ypred_test,
              sd_train = sd_train,
              sd_test = sd_test,
              Yerr_train = Yerr_train,
              Yerr_test = Yerr_test
  ))
  
}

pca_km_errorStats <- function(fit, nround = 2, compare_ix = NULL){
  # Calculate the error statistics from a pca_km object
  
  # fit ........... output object from pca_km
  # nround ........ decimals to round in output
  # compare_ix..... indices of the training set to calculate error stats for 

  if(is.null(compare_ix)){
    
    compare_ix <-  1:nrow(fit$Yerr_train)
    
  }
  
  else{ compare_ix <- compare_ix}
  
  
  trainMAE <- round(mean(abs(fit$Yerr_train[compare_ix, ])), nround)
  
  testMAE <- round(mean(abs(fit$Yerr_test)), nround)
  
  
  return(list(trainMAE = trainMAE,
              testMAE = testMAE))
  
}


pca_km_tsSparkPlot <- function(pca_km_obj, nr, nc, transp, maintext, yrs, obscol = 'black', predcol = 'red', ...){
  
  # Timeseries test set prediction sparkline plot (small multiples)
  
  par(mfrow = c(nr,nc), mar = c(0.1, 0.1, 0.1, 0.1), oma = c(1,0.1,5,0.1))
  
  ylim <- range(
    (pca_km_obj$Ypred_test + 2*(pca_km_obj$sd_test)),
    ( pca_km_obj$Ypred_test - 2*(pca_km_obj$sd_test)))
  
  for(i in 1:length(pca_km_obj$test_ix)){
    
    
    plot(yrs, (pca_km_obj$Y[test_ix, ])[i, ], ylim = ylim, type = 'n', axes = FALSE)
    
    rect(par("usr")[1], par("usr")[3],
         par("usr")[2], par("usr")[4],
         col = "grey90", border = "grey90") # Color
    
    lines(yrs, (pca_km_obj$Y[test_ix, ])[i, ], col = obscol, lwd = 1.5)
    
    lines(yrs, pca_km_obj$Ypred_test[i, ], col = predcol, lwd = 1.5)
    
    lines(yrs, pca_km_obj$Ypred_test[i, ] + 2*(pca_km_obj$sd_test[i, ]), col = makeTransparent(predcol,transp))
    lines(yrs, pca_km_obj$Ypred_test[i, ] - 2*(pca_km_obj$sd_test[i, ]), col = makeTransparent(predcol,transp))
    
 
    
  }
  
  mtext(maintext, side = 3, outer = TRUE, cex = 1.5, line = 1)
  reset()
  legend('topleft', col = c(obscol, predcol, makeTransparent(predcol,transp)),
         legend = c('observed', 'predicted', '+-2sd'), lty = 'solid', horiz = TRUE, bty = 'n')
  
}


# A function to identify rows with outlier data (from Google Bard, adapted to update threshold)
findAnomalies <- function(matrix, iqrThres = 1.5, madThres = 1.5, na.rm = FALSE) {
  # Check for NA values
  if (!na.rm) {
    naRows <- which(apply(matrix, 1, is.na))
    if (length(naRows) > 0) {
      return(naRows)
    }
  }

  # Check for outliers using IQR method
  iqr <- IQR(matrix, na.rm = na.rm)
  q1 <- quantile(matrix, 0.25, na.rm = na.rm)
  q3 <- quantile(matrix, 0.75, na.rm = na.rm)
  outlierRows <- which(apply(matrix, 1, function(row) {
    any(row < q1 - iqrThres * iqr | row > q3 + iqrThres * iqr)
  }))

  # Check for discontinuities
  discontinuityRows <- which(apply(matrix, 1, function(row) {
    any(diff(row) > madThres * mad(row))
  }))

  # Combine NA, outlier, and discontinuity rows
  anomalyRows <- union(union(naRows, outlierRows), discontinuityRows)

  # Return index of anomaly rows
  return(anomalyRows)
}



findAnomaliesTSLocal <- function(matrix, iqrThres = 5, madThres = 3, na.rm = FALSE) {
  # Find outlier runs in an ensemble
  
  # Check for NA values
  if (!na.rm) {
    naRows <- which(apply(matrix, 1, is.na))
    if (length(naRows) > 0) {
      return(naRows)
    }
  }

  # find the local (per column) IQR
  iqr <- apply(matrix, 2, FUN = IQR, na.rm = na.rm)
  
  q1 <- apply(matrix,2, FUN = quantile, probs = 0.25)
  q3 <- apply(matrix,2, FUN = quantile, probs = 0.75)
  
  
  outlierRows <- which(apply(matrix, 1, function(row) {
    any(row < q1 - iqrThres * iqr | row > q3 + iqrThres * iqr)
  }))

  # Check for discontinuities
  discontinuityRows <- which(apply(matrix, 1, function(row) {
    any(diff(row) > madThres * mad(row))
  }))

  # Combine NA, outlier, and discontinuity rows
  anomalyRows <- union(union(naRows, outlierRows), discontinuityRows)

  # Return index of anomaly rows
  return(anomalyRows)
}



```



```{r}
# Load data
load('data/ensemble-jules-historical-timeseries.RData')

```

```{r}
# Use only the post-1950 years

years_ix <- 101:164
years_trunc <- years[years_ix]

```


## Cumulative NBP

For each input/output set, it would be useful to have some functions which identify ensemble members which have "bad" data. Outliers, or NAs,
for example.




```{r}
cnbp_all_ens_wave00 <-  t(apply(nbp_ens_wave00, 1, FUN = cumsum))
cnbp_all_ens_wave01 <-  t(apply(nbp_ens_wave01, 1, FUN = cumsum))


```

```{r, fig.width = 5, fig.height = 7}
par(las = 1)

cnbp_all_col <- c(rep(makeTransparent('black',100), nrow(cnbp_all_ens_wave00)), rep(makeTransparent('red',100), nrow(cnbp_all_ens_wave01)) )

cnbp_all <- rbind(cnbp_all_ens_wave00, cnbp_all_ens_wave01)

matplot(years, t(cnbp_all_ens_wave00), type = 'l', lty = 'solid', col = makeTransparent('black',100), ylim = c(-130, 260), lwd =2,
        ylab = 'CNBP (GtC/yr)', main = 'Cumulative NBP')

legend('topleft', legend = c('Wave00', 'Wave01' ), col =c('black', 'red'), lty = 'solid')



#matplot(years, t(cnbp_all), type = 'l', lty = 'solid', col = cnbp_all_col, ylim = c(-130, 260), lwd =2,
#        ylab = 'CNBP (GtC/yr)', main = 'Cumulative NBP')

```
```{r, fig.width = 5, fig.height = 7}
par(las = 1)

cnbp_all_col <- c(rep(makeTransparent('black',100), nrow(cnbp_all_ens_wave00)), rep(makeTransparent('red',100), nrow(cnbp_all_ens_wave01)) )

cnbp_all <- rbind(cnbp_all_ens_wave00, cnbp_all_ens_wave01)

matplot(years, t(cnbp_all_ens_wave00), type = 'l', lty = 'solid', col = makeTransparent('black',100), ylim = c(-130, 260), lwd =2,
        ylab = 'CNBP (GtC/yr)', main = 'Cumulative NBP')

legend('topleft', legend = c('Wave00', 'Wave01' ), col =c('black', 'red'), lty = 'solid')

matlines(years, t(cnbp_all_ens_wave01), type = 'l', lty = 'solid', col = makeTransparent('red',100), lwd =2)


```


```{r}


cnbp_ens_wave00 <-  t(apply(nbp_ens_wave00[ , years_ix], 1, FUN = cumsum))
cnbp_ens_wave01 <-  t(apply(nbp_ens_wave01[ , years_ix], 1, FUN = cumsum))

```


```{r}

kill_ix_wave00 <- c(which(X[, 'f0_io'] > 0.9 | X[, 'b_wl_io'] < 0.15 ))

# remove those iqrThres times outside the IQR
kill_ix_wave01 <- findAnomaliesTSLocal(cnbp_ens_wave01, iqrThres = 5, madThres = 3)


```


```{r}

X_combine <- rbind(X[-kill_ix_wave00, ], X_wave01_train[-kill_ix_wave01, ])
Y_combine  <- rbind(cnbp_ens_wave00[-kill_ix_wave00,], cnbp_ens_wave01[-kill_ix_wave01, ])

Y_remove <- rbind(cnbp_ens_wave00[kill_ix_wave00,], cnbp_ens_wave01[kill_ix_wave01, ])

```


```{r, fig.width = 7, fig.height = 5}

par(mfrow = c(1,2))
matplot(years_trunc, t(Y_combine), type = 'l', lty = 'solid',
        col= c( rep('black', nrow(cnbp_ens_wave00[-kill_ix_wave00, ])), rep('red', nrow(cnbp_ens_wave01[-kill_ix_wave01, ]))),
        main = 'retained runs', ylab = 'CNBP (GtC/yr)', xlab = 'year')

legend('topleft', legend = c('Wave00', 'Wave01' ), col =c('black', 'red'), lty = 'solid')

matplot(years_trunc, t(Y_remove), type = 'l', lty = 'solid', ylim = c(-100, 100),
col = c(rep('black', length(kill_ix_wave00)), rep('red', length(kill_ix_wave01))),
main = 'removed runs', ylab =  'CNBP (GtC/yr)', xlab = 'year')


```


```{r}
testprop <- 0.2

nruns <- nrow(Y_combine)
ntest <- floor(nruns * testprop)
ntrain <- (nruns - ntest)
  
  
train_ix <- 1:ntrain
test_ix  <- (ntrain+1):nruns
```


```{r}
pc_km_cnbp_combine <- pca_km(X = X_combine,
                             Y_combine,
                             train_ix = train_ix, 
                             test_ix = test_ix, 
                             npctol = 0.99, 
                             scale = TRUE,
                             center = TRUE)

```


```{r, fig.width = 10, fig.height = 4}
par(mfrow = c(1,4), las = 1)
cnbp_pca_summary <- summary(pc_km_cnbp_combine$pca)
plot(1:10, cnbp_pca_summary$importance[3,1:10], type = 'b', xlab = 'PCs', ylab = 'Proportion of variance explained')
abline(h = 0.99, lty = 'dashed')

for(i in 1:3){

plot(years_trunc, pc_km_cnbp_combine$pca$rotation[,i], type = 'l', ylab = '', xlab = 'year')
  
}

pca_km_errorStats(pc_km_cnbp_combine) 

```



```{r, fig.width = 12, fig.height = 4}

par(mfrow = c(1,3), las = 1)
for(i in 1:3){
  
  rn <- range(c((pc_km_cnbp_combine$scores_em_train_mean[,i] - 2*(pc_km_cnbp_combine$scores_em_train_sd) ),
                (pc_km_cnbp_combine$scores_em_train_mean[,i] + 2*(pc_km_cnbp_combine$scores_em_train_sd) )
                ))
  
plot(pc_km_cnbp_combine$pca$x[,i], pc_km_cnbp_combine$scores_em_train_mean[,i], main = paste0('PC ',i),
     xlab = 'observed', ylab = 'predicted', ylim = rn, pty = 'n')
  
  segments(pc_km_cnbp_combine$pca$x[,i],   (pc_km_cnbp_combine$scores_em_train_mean[,i] - 2*(pc_km_cnbp_combine$scores_em_train_sd) ),
           pc_km_cnbp_combine$pca$x[,i],  (pc_km_cnbp_combine$scores_em_train_mean[,i] + 2*(pc_km_cnbp_combine$scores_em_train_sd) ),
           col = makeTransparent('black', 50)
           
           )
  
  points(pc_km_cnbp_combine$pca$x[,i], pc_km_cnbp_combine$scores_em_train_mean[,i], pch = 20, col = 'black')
  
  abline(0,1)
}


```





```{r, fig.width = 10, fig.height = 10}

pca_km_tsSparkPlot(pc_km_cnbp_combine, nr = 8, nc = 20, yrs = years_trunc, maintext = 'CNBP test predictions', transp = 100) 

```


```{r, fig.width = 10, fig.height = 7}

par(mfrow = c(1,2))


matplot(years_trunc, t(Y_combine[train_ix, ]), type = 'l', lty = 'solid', col = makeTransparent('black',100), lwd = 2,
        main = "Train",
        ylab = 'CNBP (GtC/yr)')
matlines(years_trunc, t(pc_km_cnbp_combine$Ypred_train), lty = 'solid', col = makeTransparent('red',100), lwd = 2)

legend('topleft', col = c('black', 'red'), legend = c('Observed', 'Predicted'), lty = 'solid')


matplot(years_trunc, t(Y_combine[test_ix, ]), type = 'l', lty = 'solid', col = makeTransparent('black',100), lwd = 2,
        main = 'Test',
        ylab = 'CNBP (GtC/yr)')
matlines(years_trunc, t(pc_km_cnbp_combine$Ypred_test), lty = 'solid', col = makeTransparent('red',100), lwd = 2)
legend('topleft', col = c('black', 'red'), legend = c('Observed', 'Predicted'), lty = 'solid')


```


```{r, fig.width = 6, fig.height = 6}

plot(Y_combine[train_ix, ], pc_km_cnbp_combine$Ypred_train, pch = 20, col = makeTransparent('black', 100),
     xlim = c(-50, 180), ylim = c(-50, 180))
points(Y_combine[test_ix, ], pc_km_cnbp_combine$Ypred_test, col = makeTransparent('red', 100), pch = 20)


abline(0,1)



```

## NPP


```{r, fig.width = 7, fig.height = 7}

par(mfrow = c(1,2))

npp_all_col <- c(rep(makeTransparent('black',100), nrow(npp_ens_wave00)), rep(makeTransparent('red',100), nrow(npp_ens_wave01)) )

npp_all <- rbind(npp_ens_wave00, npp_ens_wave01)

npp_anom_all <- anomalizeTSmatrix(npp_all, 1:20)


matplot(years, t(npp_all), type = 'l', lty = 'solid', col = npp_all_col, ylim = c(-10, 180), lwd =2,
        ylab = 'NPP (GtC/yr)')
matplot(years, t(npp_anom_all), type = 'l', lty = 'solid', col = npp_all_col, ylim = c(-20, 45), lwd =2,
        ylab = 'NPP anomaly (GtC/yr)')





```


```{r}

kill_ix_wave00 <- c(which(X[, 'f0_io'] > 0.9 | X[, 'b_wl_io'] < 0.15 ))

# remove those iqrThres times outside the IQR
kill_ix_wave01 <- findAnomaliesTSLocal(npp_ens_wave01[, years_ix], iqrThres = 5, madThres = 3)


```


```{r}

X_combine <- rbind(X[-kill_ix_wave00, ], X_wave01_train[-kill_ix_wave01, ])

# WARNING - if nothing is removed, nothing is combined
Y_combine  <- rbind(npp_ens_wave00[-kill_ix_wave00, years_ix], npp_ens_wave01[-kill_ix_wave01, years_ix])

Y_remove <- rbind(npp_ens_wave00[kill_ix_wave00, years_ix], npp_ens_wave01[kill_ix_wave01, years_ix])

```


```{r, fig.width = 7, fig.height = 7}

par(mfrow = c(1,2))
matplot(years_trunc, t(Y_combine), type = 'l', lty = 'solid', col = makeTransparent('black',100),
        lwd = 2, main = 'retained runs',
        ylab = 'NPP (GtC/yr)')
matplot(years_trunc, t(Y_remove), type = 'l', lty = 'solid',
col = c(rep('black', length(kill_ix_wave00)), rep('red', length(kill_ix_wave01))),
main = 'removed runs', ylab = 'NPP (GtC/yr)')


```



```{r}
testprop <- 0.2

nruns <- nrow(Y_combine)
ntest <- floor(nruns * testprop)
ntrain <- (nruns - ntest)
  
  
train_ix <- 1:ntrain
test_ix  <- (ntrain+1):nruns
```


```{r}
pc_km_npp_combine <- pca_km(X = X_combine,
                             Y_combine,
                             train_ix = train_ix, 
                             test_ix = test_ix, 
                             npctol = 0.99, 
                             scale = TRUE,
                             center = TRUE)

```


```{r, fig.width = 10, fig.height = 10}

pca_km_tsSparkPlot(pc_km_npp_combine, nr = 8, nc = 20, yrs = years_trunc, maintext = 'NPP test predictions', transp = 100) 

```

## NPP anomaly
```{r}

npp_anom_ens_wave00 <- anomalizeTSmatrix(npp_ens_wave00[, years_ix], 1:20)
npp_anom_ens_wave01 <- anomalizeTSmatrix(npp_ens_wave01[, years_ix], 1:20)
  

kill_ix_wave00 <- c(which(X[, 'f0_io'] > 0.9 | X[, 'b_wl_io'] < 0.15 ))

# remove those iqrThres times outside the IQR
kill_ix_wave01 <- findAnomaliesTSLocal(npp_anom_ens_wave01, iqrThres = 5, madThres = 3)


```


```{r}

X_combine <- rbind(X[-kill_ix_wave00, ], X_wave01_train[-kill_ix_wave01, ])

# WARNING - if nothing is removed, nothing is combined
Y_combine  <- rbind(npp_anom_ens_wave00[-kill_ix_wave00, ], npp_anom_ens_wave01[-kill_ix_wave01, ])

Y_remove <- rbind(npp_anom_ens_wave00[kill_ix_wave00, ], npp_anom_ens_wave01[kill_ix_wave01, ])

```


```{r}

par(mfrow = c(1,2))
matplot(years_trunc, t(Y_combine), type = 'l', lty = 'solid', col = 'black', main = 'retained runs')
matplot(years_trunc, t(Y_remove), type = 'l', lty = 'solid',
col = c(rep('black', length(kill_ix_wave00)), rep('red', length(kill_ix_wave01))),
main = 'removed runs')


```



```{r}
testprop <- 0.2

nruns <- nrow(Y_combine)
ntest <- floor(nruns * testprop)
ntrain <- (nruns - ntest)
  
  
train_ix <- 1:ntrain
test_ix  <- (ntrain+1):nruns
```


```{r}
pc_km_npp_anom_combine <- pca_km(X = X_combine,
                             Y_combine,
                             train_ix = train_ix, 
                             test_ix = test_ix, 
                             npctol = 0.99, 
                             scale = TRUE,
                             center = TRUE)

```


```{r, fig.width = 10, fig.height = 10}

pca_km_tsSparkPlot(pc_km_npp_anom_combine, nr = 8, nc = 20, yrs = years_trunc, maintext = 'NPP anomaly test predictions', transp = 100) 

```







## Run PC emulator for all outputs

```{r}
varnames <- c("baresoilFrac_lnd_mean", 
                "c3PftFrac_lnd_mean",
                "c4PftFrac_lnd_mean", 
                #"cnbp",
                "cSoil",
                "cVeg", 
                "fHarvest_lnd_sum", 
                "fLuc_lnd_sum", 
                "lai_lnd_mean",         
                "nbp",
                "npp",  
                "rh_lnd_sum",
                "shrubFrac_lnd_mean", 
                "treeFrac_lnd_mean")

```



```{r, fig.width = 10, fig.height = 10}

kill_ix_wave00 <- c(which(X[, 'f0_io'] > 0.9 | X[, 'b_wl_io'] < 0.15 ))
testprop <- 0.2

for(i in 1:length(varnames)){
  
  varname <- varnames[i]
  
  ens_wave00 <- get(paste0(varname, '_ens_wave00'))[, years_ix]
  ens_wave01 <- get(paste0(varname, '_ens_wave01'))[ , years_ix]

  
# remove those iqrThres times outside the IQR
  kill_ix_wave01 <- findAnomaliesTSLocal(ens_wave01, iqrThres = 5, madThres = 3)
  
  if(identical(kill_ix_wave01, integer(0))){
    
    X_combine <- rbind(X[-kill_ix_wave00, ], X_wave01_train)
    Y_combine  <- rbind(ens_wave00[-kill_ix_wave00, ], ens_wave01)
    Y_remove <- rbind(ens_wave00[kill_ix_wave00, ])
  
  
  }
  else{
    X_combine <- rbind(X[-kill_ix_wave00, ], X_wave01_train[-kill_ix_wave01, ])
    Y_combine  <- rbind(ens_wave00[-kill_ix_wave00, ], ens_wave01[-kill_ix_wave01, ])
    Y_remove <- rbind(ens_wave00[kill_ix_wave00, ], ens_wave01[kill_ix_wave01, ])
  }


  #par(mfrow = c(1,2), oma = c(0.1, 0.1, 5, 0.1))
  #matplot(years_trunc, t(Y_combine), type = 'l', lty = 'solid', col = 'black', main = 'retained runs')
  #matplot(years_trunc, t(Y_remove), type = 'l', lty = 'solid',
  #col = c(rep('black', length(kill_ix_wave00)), rep('red', length(kill_ix_wave01))),
  #main = 'removed runs')
  #mtext(side = 3, text = varname, outer = TRUE, line = 3, cex = 2)
  
  outname <- paste0('pc_km_', varname)

  nruns <- nrow(Y_combine)
  ntest <- floor(nruns * testprop)
  ntrain <- (nruns - ntest)
  
  train_ix <- 1:ntrain
  test_ix  <- (ntrain+1):nruns
  
  pc_km <- pca_km(X = X_combine,
                             Y_combine,
                             train_ix = train_ix, 
                             test_ix = test_ix, 
                             npctol = 0.99, 
                             scale = TRUE,
                             center = TRUE)
  
  assign(outname, pc_km)
  
  pca_km_tsSparkPlot(get(outname), nr = 8, nc = 20, yrs = years_trunc, maintext = varname, transp = 100) 
  
}


```



